---
title: "Homework 2"
author: "Noah Duran"
date: "Assigned: 29 Oct"
output: 
  html_document:
    theme: paper
    highlight: tango
---

## To complete this assignment, follow these steps:

1. Download the `HW2.Rmd` file from Canvas.
2. Open `HW2.Rmd` in RStudio.
3. Replace the "your name here" text in the `author:` field with your own name.
4. Supply your solutions to the homework by editing `HW2.Rmd`.
5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, save the html file as `HW2_YourAndrewIDHere.html`, and submit on Canvas.  (YourAndrewIDHere should be changed to your Andrew ID, of course.)

## Homework tips:

1. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.
2. Run your code in the Console and Knit HTML frequently to check for errors.
3. You may find it easier to solve a problem by interacting only with the Console at first. 

## Homework 2 outline

## Problem 1: Simple Boolean operations

> Tip: To check that your answer makes sense, you may want to try it out in the console with various choices of values for the variable `x`.   

## (a) Checking equality.

Define a variable `x` and write a Boolean expression that evaluates to `TRUE` if the variable `x` is equal to `9452` (the numeric value).  

```{r question 1a}
# Insert your Boolean expression here
x <- 9452

print(x == 9452)

```

## (b) Checking inequality.

Given any variable `x`, write a Boolean expression that evaluates to `TRUE` if the variable `x` is *not* `NA` (i.e., is not missing).

```{r question 1b}
# Insert your Boolean expression here
x <- 123
!is.na(x)

```

## (c) Checking if a number is in a given range. 

Given a (possibly negative) number `x`, write a Boolean expression that returns `TRUE` if and only if `x` is less than `-6` OR greater than `11`.  

```{r question 1c}
# Insert your Boolean expression here
x <- 14
print(x < -6 | x > 11)

```

## Problem 2: Multiple Boolean operations

## (a) select multiple values from a vector.  

A numeric vector v is defined in the code chunk below. Create a second vector v1 using a Boolean expression with only those values between 5 and 10 and secondly determine the length of this new vector.

```{r question 2a}
# numeric vector definition
v <- c(6, 10, 2, 11, 7, 12, 4, 5, 1, 1, 9, 8, 11, 5, 6, 7, 1, 2, 10, 12)
# Insert your Boolean expression here to calculate v1
v1 <- v[v > 5 & v < 10] 
#if the values between 5 and 10 should be inclusive the code would be 
# v1 <- v[v >= 5 & v <= 10]
# determine the length of vector v1
length(v1)
```

## (b) Using `all()`

Two people were asked to give their preferences between two options: [Facebook, Twitter], [Firefox, Chrome], [Mac, PC], [Spring, Fall].  Their results are given below.

```{r question 2b}
alice_prefs <- c("Twitter", "Chrome", "Mac", "Spring")
bob_prefs <- c("Facebook", "Chrome", "PC", "Fall")
```

Use the `all()` function to determine if the two people have identical preferences. (Your code should output a single Boolean value, either `TRUE` or `FALSE`)

```{r question 2b-2}
# Edit me
print(all(alice_prefs == bob_prefs))
```

## (c) Using `any()` 

Use the `any()` function to determine if the two people have any preferences in common.  (Your code should output a single Boolean value, either `TRUE` or `FALSE`)

```{r question 2c}
# Edit me
print(any(alice_prefs == bob_prefs))

```

## (d) Missing values.

Let `age` be the vector defined below.

```{r question 2d}
age <- c(18, 17, 25, 34, NA, 45, NA, 45, 50, NA, 19, 18, 22)
```

Write a Boolean expression that checks whether each entry of `age` is missing (recall missing values are denoted by `NA`).  Your expression should return a Boolean vector having the same length as the `age` vector.

```{r question 2d end}
# Edit me
is.na(age)
```

## Problem 3: Referencing vector elements

## (a) `which()` practice

Write code that returns the indices of the `age` vector that are missing.

```{r question 3a}
# Edit me
which(is.na(age))
```

## (b) Getting non-missing values

Write code that uses negative indexes and your solution from (a) to return only the values of `age` that are *not* missing. (i.e., your code should result in a vector with elements: 18, 17, 25, 34, 45, 45, 50, 19, 18, 22


```{r question 3b}
# Edit me
negative.index.age <- which(is.na(age))
age[-negative.index.age]
```

## (c)  A more direct way of getting non-missing values

Using the negation operator `!` and the `is.na()` function, write an expression that returns only the values of `age` that are *not* missing.

```{r question 3c}
# Edit me
age[!is.na(age)]
```

## (d) More `which()` practice

The next few questions use the built-in `cars` data set.

```{r question 3d}
speed <- cars$speed
dist <- cars$dist
```

Write code to determine which of the entries in cars has a stopping distance (dist) of 40 feet or more.

```{r question 3d-2}
# Edit me
dist.over40 <- cars$dist >= 40
print(dist.over40)

```

## (e) `which.min`, `which.max` practice

Use the `which.min()` function to figure out which car had the *shortest* stopping distance. (Your code should return the car's index.)

```{r question 3e}
# Edit me
which.min(dist)
```

## (f) More practice 

Use the `which.max()` function to figure out the *speed* of the car that had the *longest* stopping distance.  (Your code should return the car's speed.)

```{r question 3f}
# Edit me
which.max(dist)
```

## Problem 4: Data frame basics

## (a) Importing data

Use the `read.csv()` function to import the `survey.csv` dataset into a variable called `survey`.

```{r question 4a}
# Edit me
survey <- read.csv("~/Desktop/Masters Degree/R Programming/datasets/survey.csv", header=TRUE, sep=",", stringsAsFactors = TRUE)
```

**Tip**: The csv data file called `survey.csv` is available on Canvas.  You may download the file to a directory on your machine then import it.

## (b) `$` notation

Use the `$` operator to select and return the contents of the studyhours column from the `survey` dataset

```{r question 4b}
# Edit me
print(survey$StudyHours)
```

## (c) [,] notation

Repeat part (b) using `[]` notation.  i.e., Use `[]` notation to select the studyhours column from the `survey` data by position (i.e., obtain this column by using the proper column number)

```{r question 4c}
# Edit me
print(survey[, 8])
```

## (d) [[]] notation

Repeat part (c) with [[]] notation.

```{r question 4d}
# Edit me
print(survey[["StudyHours"]])
```

## (e) `subset()` practice

Use the `subset()` function to select all the survey data on Program and OperatingSystem for respondents whose Rexperience is "Never used" *OR* who studied 6 or less hours per week.

```{r question 4e}
# Edit me
subset(survey, 
       select=c("OperatingSystem", "Program"), 
       Rexperience == "Never used" | StudyHours <= 6)

```

## Problem 5: Data summaries and inline code practice.

## (a) Bar graph

Create a bar graph of respondents' Rexperience.

```{r question 5a, fig.align='center', fig.width=7, fig.height=4}
# Edit me
plot(survey$Rexperience,
    main = "Student Experience with R",
    xlab = "Experience Level",
    ylab = "Number of responses")
```

## (b) Inline code practice

Replace all occurrences of ???? in the paragraph below with inline code supplying the appropriate information.

> Of the `r ncol(survey)` respondents, `r sum(survey[["Program"]] != "MISM")` were NOT from the MISM program.  We found that `r mean(survey[["OperatingSystem"]] == "Mac OS") * 100`% of the all students in the class use the Mac OS X operating system.  `r mean(survey[["Rexperience"]][survey[["Program"]] == "MISM"] == "Basic competence") * 100`% of the MISM students report having Basic competence in R.


